Diversification of Australian blind snakes

### 2022-09-26 manuscript.Rmd
update results biogeobears

@@

@@

@@

@@

>>>>

The aridification on continental Australia over the past 20 million years resulted in a diversity of habitats and spurred multiple radiations of flora and fauna across the continent, many of which are endemic [@byrneBirth2008; @melvilleEvolutionary2011]. The arid zone makes up the majority of the Australian landscape and provides habitat to ~43% of all Australia’s terrestrial reptiles. Squamate reptiles (lizards and snakes) represent the largest vertebrate diversity in Australia with over 1000 species (~800 lizards and ~200 snakes), many of which have undergone rapid diversification in the last 10 – 6 myr [@brennanMass2017; @oliverPhylogenetic2017]. For example, some Australian squamate clades like dragons [@melvilleIntegrating2019], sphenomorphine skinks [@raboskyAnalysis2014], geckos [@ashmanDiversification2018; @brennanMass2017] and elapid snakes [@leeDiversification2016; @grundlerTrophic2014] show increased rates of diversification coincidentally with the onset of aridification. However, these studies did not account for heterogeneity of the arid zone [@pepperLife2021].

Roughly a quarter of the snake species diversity in Australia is represented by blindsnakes of the genus _Anilios_, which comprises 48 described species [@uetzQuarter2021]. All _Anilios_ species are found in Australia except for _A. eryncinus_, which is only found in Papua New Guinea. While morphologically conserved and difficult to distinguished, extensive phylogenetic studies and several efforts to revisit specimens revealed the true species diversity is underestimated [@marinHidden2013; @pyronSystematics2014; @ellisTyphlopid2019]. The diversification is reported to have originated from a single colonisation event from Asia between 19-39 mya [@vidalBlindsnake2010; @marinTracing2013], before the onset of aridification over the past 20 million years [@byrneBirth2008]. However, the study did not account for alternative biogeographic areas (e.g. WWF Ecoregions) and models, or uncertainty in phylogenetic relationships and divergence times. Despite representing a substantial portion of snake biodiversity there has been few investigations into the macroevolutionary dynamics of blind snakes. Recent phylogenomic tree and monophyly of the genus makes the Australian _Anilios_ well suited to test hypotheses about how lineages diversified with dramatic ecological changes because they are geographically restricted to the continent and can be found in almost every biome [citefirstchapter?].

<!-- Environmental correlates of diversification -->
<!-- agamid lizards [@harmonTempo2003], passerine birds [@ricklefsGlobal2006], amphibians [@rollandContribution2019], rosid plants [@sunRecent2020]).  -->
A variety of ecological correlates have been associated with diversification processes of a clade. Among others, studies on the influence of climate (e.g. temperature, precipitation, and latitude) [@ricklefsGlobal2006; @mayhewBiodiversity2012; @condamineMacroevolutionary2013; @cantalapiedraDietary2014; @condamineClimate2019; @rollandContribution2019; @sunRecent2020]], elevation [@esquerreHow2019; @boschmanMountain2021], biological interactions [@ramirezAsynchronous2011; @druryEstimating2016; @brennanPhylogenomicsMonitorLizards2021], CO<sub>2</sub> [@kergoatOpposite2018], microhabitats [@bars-closelDiversification2017] and aridification [@raboskyExceptional2007; @pepperGenetic2011; @deepakAridification2018; @catulloAridification2014] have been documented. In Australia, shifts in diversification rates owing to changes in temperature and aridity has been reported in skinks [@raboskyExceptional2007], geckos [@fujitaDiversification2010; @pepperGenetic2011; @pepperPalaeoclimate2011; @oliverIndependent2014; @oliverContrasting2014], frogs [@catulloAridification2014; @vidal-garciaConvergent2015], saltbush [@kadereitMolecular2010], *Acacia* [@millerComparative2013], insects [@owenHow2017; @toussaintUnveiling2015], and other plants [reviewed in @crispHow2013].

There are several links between temperature and diversification rates of reptiles. First, warm temperature may increase resource availability, which in turn increase ecological niche in general [@clarkeClimate2006]. Second, spatial and temporal heterogeneity in temperature may lead to climatic niche divergence, such as local adaptations in different geographic areas [@pepperGeodiversity2013], and change in activity time of prey/predators [@santosTemperature2011; @lawsonLatitudinal2014]. Third, change in climate change (e.g. glaciation) may cause fragmentation of continuous habitats leading to isolated populations, which may lead to divergence [@oliverRocks2019]. Fourth, ectotherms are subjected to higher metabolic rates in warm climates (i.e. Kleiber's law), which may increase mutation rates and ultimately greater speciation rates [@allenKinetic2006; @makarievaMean2008]. Finally, aridification of Australia may offer new ecological niche for species to exploit, which have been shown to result in deep divergences in many arid-adapted [@byrneBirth2008; @pepperGenetic2011; @marinTracing2013].

The influence of temperature on diversification rates has been empirically tested with birth-death models [@condamineMacroevolutionary2013; @kergoatOpposite2018; @condamineClimate2019; @rollandContribution2019], these temperature data are reconstructed (interpolated) based on conversion of concentration of oxygen isotopes [@condamineMacroevolutionary2013]. However, the relationship between diversification rates and aridity are attributed to some proxy such as the the timing of aridification or global temperature rather than quantifying the extent of aridity itself [@dornStrange2014; @deepakAridification2018; @catulloAridification2014; @fujitaDiversification2010; @pepperGenetic2011; @pepperPalaeoclimate2011; @oliverIndependent2014; @oliverContrasting2014; @owenHow2017; @crispHow2013; @toussaintUnveiling2015]. As such, no study has been able to quantitatively assess aridity as a continuous variable in environment-dependent models and compare it to other ecological processes. Recent development in geographic-based methods such as `gen3sis` modeling engine [@hagenGen3sis2021] enables us to simulate climatic, geological, and ecological factors that are spatially and temporally specific to certain areas. In practice, these outputs from `gen3sis` provide paleotemperature and aridity data specific to Australia (instead of relying on global paleoclimate data) in the time span of interest (between the 45 mya and the present). These datasets will allow us to empirically test the influence of localised ecological processes on diversification rates over time.

In this study, we use the Australian radiation of blindsnakes as a system to investigate the diversification of morphologically conserved squamates in arid and non-arid zones by testing hypotheses about the processes that promote speciation. Applying the cutting edge localised geographic-based paleoenvironment reconstruction, we evaluated (1) the model fit performance of localised versus globalised temperature on the diversification of Australian blindsnake radiation, (2) the model fit between different ecological processes on diversification rates, and (3) What is the ancestral range of Australian blindsnakes? (4) Are there differences in the rates of speciation between arid and non-arid zones?

@@

## Reconstruction of Australian paleo climate 
<!-- Consult Alex Skeels -->
Two paleographic digital elevation models (DEMs) providing paleotopography at resolution of 1&deg; x 1&deg; and 5 myr resolution  were obtained from Scotese (2021) [@scoteseAtlas2021] and Straume et al (2020) [@straumeGlobal2020]. We projected the paleoenvironmental reconstructions into an equal-area Behrmann projection of 110km x 110km following methods described in Hagen et al (2021) @hagenGen3sis2021. From the projected data we extracted temperature and aridity values across all grid cells that fall within Sahul by manually drawing a polygon that encompassed the continent from 40 mya to the present-day (\@ref(fig:fig-aus-region).

## Phylogenetic hypothesis
<!-- Based on our phylogenetic tree -->
We employ the recent time-calibrated phylogenomic tree of Australian blindsnake with outgroups. This tree includes __ taxa representing __/48 described species of _Anilios_. However, because we our phylogenetic data suggest some cryptic species, we adjust our taxon sampling fraction to x/52 species.

## Biogeography

To evaluate the ancestral biome of blindsnakes in Australia, we used the R package `BioGeoBEARS` 1.1.1 [@nicholas_j_matzke_2018_1478250]. BioGeoBEARS provides a flexible likelihood-based framework that models how the branches in the phylogeny (representing extant taxa and their ancestors) have shifted among a set of discrete biomes over time. These shifts in biomes can occur within a single branch (anagenetic events including processes like dispersal, extinction, or switching to adjacent biomes) or at the time of branching (cladogenetic events involving sympatric, vicariant and founder-event speciation). The fundamental inputs for the software include a time-calibrated phylogeny (described above), information about the biome/range each species are found, and a historical biogeographic model. We divided Australo-Papua into nine discrete biomes and assigned each Australian species to specific biome(s) in which the based on occurrence records from the online database [Atlas of the Living Australia](doi.ala.org.au/doi/10.26197/ala.d92678b1-ad2d-437b-9457-9f52737ba003) (see Supplementary information S2 for how data was compiled). 
Biome designations were based on the World Wildlife Fund's (WWF) Terrestrial Ecoregions of the World dataset [@olsonTerrestrial2001] and followed the nomenclature of previous Austrlia-specific studies [@cardilloPhylogeny2017; @pavon-vazquezComprehensive2021].

The phylogeny and distribution information were then applied to widely used historical biogeographical methods including Dispersal-Vicariance Analysis (DIVA; [@ronquistDispersalVicariance1997]), and likelihood implementations of Dispersal-Extinction-Cladogenesis (DEC; [@reeMaximum2008]) or DECLIKE, and BayArea [@landisBayesian2013] or BAYAREALIKE [@matzkeBioGeoBEARS2013]. Additionally, we tested these models with distance matrix as a free parameter (+X). 
We then compared these models and selected the best models based on AICc (corrected Akaike Information Criterion) scores. The best-fitting model was then used to calculate the probability of the ancestral states by using the BioGeoBEARS stochastic mapping function (`N` runs).

<!-- Include environments on range evolution ?? -->

The best-fitting model was then used to calculate the probability of the ancestral states by using the BioGeoBEARS stochastic mapping function (`N` runs).

We tested several biogeographic models including  and their modifiers (See supplementary).

<!-- SUPPLEMENT TABLE WITH OCCURENCE RECORDS AND BIOME MAP -->

## Diversification analyses

To test whether and how diversification rates correlated with climate and aridity through time, we fitted several time-variable likelihood diversification models using `RPANDA` [@morlonRPANDA2016] to the _Anilios_ phylogeny. These models will be a combination of pure-birth and birth-death models with different speciation/extinction rates specified as constant, linearly, or exponentially correlated with paleoenvironment. We fitted different combinations of pure-birth and birth-death models with speciation and/or extinction rates being constant or otherwise varying linearly or exponentially with the following environment variables: 1 global paleo temperature, 3 Australian paleo temperature, 3 Australian paleo aridity, and 3 Australian paleo proportion arid area. Missing taxa is accounted for by fractioning function in RPANDA. We assessed model fit with corrected Akaike Information Criteria (AICc) scores.

- Initial lamba parameters were changed for mean_sco

## Biogeographic-dependent 
To test whether diversification rates differs between arid and non-arid zones test for biogeographic-depdendent diversification using geographic state and hidden state speciation-extinction model (GeoHiSSE) analysis with different parameters [@goldbergPhylogenetic2011].

MiSSE https://www.biorxiv.org/content/10.1101/2021.11.02.466977v1

@@

## Paleoclimate conditions

Paleoclimate reconstructions of aridity differed slightly between the two methods \@ref(fig:fig-paleoclimate-compare)a. Both aridity index shows increase in aridity post 10 mya, consistent with aridification of the continent [@byrneBirth2008]. Temperature for Australia differed between the three methods \@ref(fig:fig-paleoclimate-compare)b. Reconstructed temperature values from Australia were higher than global temperature and stayed relatively constant compared to global temperature that shows slight decrease.

## Biogeographic history

We found that the best fitting model is DEC plus jump dispersal and distance matrix (DEC+j+x). 
Likelihood ratio test (LRT) indicates that given the number of extra free parameters for the complex model, it is a statistically significant improvement over the simpler DEC, DEC+j, and DEC+x models (X^2^, DF = 2, D-statistic = 26.7, p-value < 0.001).

- Topography dyanmics [@brazModelling2021]
- @marinTracing2013 concluded that Australian blindsnakes, like several other Australian squamates, originated from tropical-mesic origins and expanded into the arid zone since the mid-Miocene [@chappleParallel2004;@melvilleEvolutionary2011]. 
- Range expansion was important for diversification ...

Recent biogeographic reconstruction recovered India as the origin for superfamily Typhlopoidea (Gerrhopilidae, Xenotyphlopidae and Typhlopidae) [@kleinEvolution2021].

## Diversification
The best fitting model among constant, time-dependent, temperature-dependent, and aridity dependent models is 
This model indicate that X associate with speciation (alpha = ) and .....

Fit with Australia specific mean temperature receives better support than model with global temperature and exponential dependency of speciation rate with time.

- Solomon Island 40 MYA [@hallCenozoic2002]

@@

## Localised climate model outperforms global models

## Ancestral biome reconstruction
- Consistent with Marin et al.
- Timing is similar to other squamates arrival

## Drivers of diversifcation 
-

@@

# Tables

Table BioGeoBEARS. BioGeoBEARS model estimates and model AICc scores. Lower AICc scores indicate model is preferred. + Log likelihood
Table RPANDA. Diversification model fit estimates and AICc score of model fit

# Figure legends

Fig. Ancestral biome: Stochastic map of Ancestral biome from Australo-Papuan blind snakes
Fig. Paleoclimate: Reconstructed paleoclimate values from different data sets. Mean temperature data. Aridity index based on precipitation threshold. Abbreviations Pli = Paleocene and Ple = Pleistocene. 
Fig. Speciation rates + phylogeny over time + lineages through time
